Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\

### 2025-05-27 paper/main.tex
autoencoder formatting issues
Updates from Overleaf

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

@@

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{float}

@@

\author{
\IEEEauthorblockN{Nuno Batista}
\IEEEauthorblockA{\textit{Department of Informatics Engineering} \\
\textit{Faculty of Sciences and Technology, University of Coimbra}\\
Coimbra, Portugal \\
\href{mailto:nunomarquesbatista@gmail.com}{nunomarquesbatista@gmail.com}}

\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

@@

@@

\subsection{Background and Motivation}
Advances in medical imaging have significantly improved 
disease diagnosis and treatment planning. For conditions 
like colorectal cancer, early detection of polyps through 
colonoscopy image analysis is critical for reducing mortality~\cite{estevaGuideDeepLearning2019a}. 
Deep learning techniques, especially autoencoders, are widely 
used to extract compressed, informative features from 
high-dimensional images for classification and segmentation 
tasks~\cite{bengioLearningDeepArchitecturesa}. However, classical neural networks may struggle 
to capture intricate correlations in complex medical data.

@@

The model is evaluated and compared to classical benchmarks on three 
different datasets: 
\begin{enumerate}
    \item A synthetic dataset of polyp images, generated to simulate realistic medical imaging scenarios.
    \item Real image patches extracted from the CVC-ClinicDB dataset, a well-known benchmark for polyp detection.
    \item A reduced version of the MNIST dataset, containing only the digits 0 and 1, suitable for binary classification tasks.
\end{enumerate}

@@

%============================================
% BACKGROUND 
%============================================
\section{Background}

@@

A diagram of a typical reservoir computing architecture is shown in `Fig.-\ref{fig:reservoir_architecture}'.

@@

\subsection{Reservoir Computing with Neutral Atoms}
As quantum Reservoir Computing (QRC) leverages the complex 
dynamics of quantum systems to process information, 
extending the classical reservoir computing 
paradigm into the quantum domain. Neutral atom platforms, 
particularly those utilizing Rydberg states, 
have emerged as promising candidates for implementing 
QRC due to their scalability and controllable interactions.

@@

The dynamics of such neutral atom systems can be 
described by the Rydberg Hamiltonian, which captures 
the essential physics of laser-driven interactions 
among atoms in Rydberg states. Following Kornjača et al.~\cite{kornjavcaLargescaleQuantumReservoir2024}
The Hamiltonian for a system of neutral-atoms is given by:

\begin{equation}
    \begin{aligned}
        H(t) &= \dfrac{\Omega(t)}{2}\sum_j \left(\ket{g_j}\bra{r_j}+\ket{r_j}\bra{g_j}\right) \\
             &{{\quad}} + \sum_{j<k}V_{jk}n_jn_k - \sum_j \left[\Delta_{\mathrm{g}}(t) + \alpha_j\Delta_{\mathrm{l}}(t)\right] n_j,
    \end{aligned}
    \label{eq:rydberg_hamiltonian}
\end{equation}

where \( \Omega(t) \) is the global Rabi drive amplitude between a 
ground state \( |g_j\rangle \) and a highly-excited Rydberg state 
\( |r_j\rangle \) of an atom,
\( n_j = |r_j\rangle \langle r_j| \), 
\( V_{jk} = C/\|r_j - r_k\|^6 \) describes the van der Waals interactions
between atoms, and the detuning is split into a global term 
\( \Delta_{\mathrm{g}}(t) \)
and a site-dependent term \( \Delta_{\mathrm{l}}(t) \), with site 
modulation \( \alpha_j \in [0, 1] \).

@@

Autoencoders can capture complex, nonlinear 
relationships in the data, making them suitable for tasks 
like image compression, denoising, and anomaly 
detection~\cite{hintonReducingDimensionalityData2006, estevaGuideDeepLearning2019a}.

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}
Our proposed pipeline integrates classical and quantum 
components for dimensionality reduction and classification. 
A classical autoencoder compresses high-dimensional image 
data into a lower-dimensional latent space. This latent 
representation is encoded into a Rydberg atom chain, 
serving as input to a quantum reservoir. The reservoir's 
dynamics, governed by the Rydberg Hamiltonian, produce 
quantum embeddings through measurements of quantum observables.

@@

The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

@@

These aspects hinder the direct application of 
gradient-based optimization methods across the 
quantum-classical boundary, obstructing end-to-end 
training of hybrid models.

@@

The readout is performed by measuring specific observables,
such as:
\begin{itemize}
    \item \textbf{Single-atom Measurements}: 
    \( \langle Z_i \rangle \) for each atom \( i \), where \( Z_i \) is the Pauli-Z operator.
    \item \textbf{Two-atom Correlations}: 
    \( \langle Z_i Z_j \rangle \) for pairs of atoms \( (i, j) \), capturing correlations between their states.
    \item \textbf{Three-atom Correlations}:
    \( \langle Z_i Z_j Z_k \rangle \) for triplets of atoms \( (i, j, k) \), providing higher-order correlations.
\end{itemize}

@@

The autoencoder was trained to minimize reconstruction 
loss using the Adam optimizer with learning rates 
between 0.001—0.01 and weight decay for regularization.

@@

\subsection{Performance Metrics}
We evaluated the performance of our model using the following metrics:
\begin{itemize}
    \item \textbf{Classification Accuracy:} 
    The percentage of correctly classified instances in the test set.

    \item \textbf{F1 Score:} 
    The harmonic mean of precision and recall, providing a balanced measure of classification performance.

    \item \textbf{Confusion Matrix:} 
    A matrix summarizing the classification results, showing true positives, false positives, true negatives, and false negatives.

\end{itemize}

@@

>>>>

\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

@@

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{float}
\usepackage{inputenc}

@@

\author{\IEEEauthorblockN{Anonymous Authors}}
% \author{
% \IEEEauthorblockN{Nuno Batista}
% \IEEEauthorblockA{\textit{Department of Informatics Engineering} \\
% \textit{Faculty of Sciences and Technology, University of Coimbra}\\
% Coimbra, Portugal \\
% \href{mailto:nunomarquesbatista@gmail.com}{nunomarquesbatista@gmail.com}}

% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address or ORCID}
% }

@@

%ADD SPACE TO BREATHE

@@

\subsection{Background and Motivation}
Advances in medical imaging have significantly improved 
disease diagnosis and treatment planning. For conditions 
like colorectal cancer, early detection of polyps through 
colonoscopy image analysis is critical for reducing mortality~\cite{estevaGuideDeepLearning2019}. 
Deep learning techniques, especially autoencoders, are widely 
used to extract compressed, informative features from 
high-dimensional images for classification and segmentation 
tasks~\cite{bengioLearningDeepArchitectures}. However, classical neural networks may struggle 
to capture intricate correlations in complex medical data.

@@

@@

%============================================
% BACKGROUND 
%============================================
\section{Background and Related Work}

@@

A diagram of a typical Reservoir Computing architecture is shown in `Fig.-\ref{fig:reservoir_architecture}'.

@@

\subsection{Reservoir Computing with Neutral Atoms}
As Quantum Reservoir Computing (QRC) leverages the complex 
dynamics of quantum systems to process information, 
extending the Classical Reservoir Computing 
paradigm into the quantum domain. Neutral atom platforms, 
particularly those utilizing Rydberg states, 
have emerged as promising candidates for implementing 
QRC due to their scalability and controllable interactions.

@@

The dynamics of such neutral atom systems can be 
described by the Rydberg Hamiltonian, which captures 
the essential physics of laser-driven interactions 
among atoms in Rydberg states. Following Kornjača et al.~\cite{kornjavcaLargescaleQuantumReservoir2024}
the Hamiltonian for a system of neutral-atoms is given by:

\begin{equation}
    \begin{aligned}
        H(t) &= \dfrac{\Omega(t)}{2}\sum_j \left(\ket{g_j}\bra{r_j}+\ket{r_j}\bra{g_j}\right) \\
             &{{\quad}} + \sum_{j<k}V_{jk}n_jn_k - \sum_j \left[\Delta_{\mathrm{g}}(t) + \alpha_j\Delta_{\mathrm{l}}(t)\right] n_j,
    \end{aligned}
    \label{eq:rydberg_hamiltonian}
\end{equation}
where \( \Omega(t) \) is the global Rabi drive amplitude between a 
ground state \( |g_j\rangle \) and a highly-excited Rydberg state 
\( |r_j\rangle \) of an atom,
\( n_j = |r_j\rangle \langle r_j| \), 
\( V_{jk} = C/\|r_j - r_k\|^6 \) describes the van der Waals interactions
between atoms, and the detuning is split into a global term 
\( \Delta_{\mathrm{g}}(t) \)
and a site-dependent term \( \Delta_{\mathrm{l}}(t) \), with site 
modulation \( \alpha_j \in [0, 1] \).

@@

Autoencoders can capture complex, nonlinear 
relationships in the data, making them suitable for tasks 
like image compression, denoising, and anomaly 
detection~\cite{hintonReducingDimensionalityData2006, estevaGuideDeepLearning2019}.

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}

Our proposed pipeline integrates classical and quantum 
components for dimensionality reduction and classification. 
A classical autoencoder compresses high-dimensional image 
data into a lower-dimensional latent space. This latent 
representation is encoded into a Rydberg atom chain, 
serving as input to a quantum reservoir. The reservoir's 
dynamics, governed by the Rydberg Hamiltonian, produce 
quantum embeddings through measurements of quantum observables.
The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

@@

@@

These aspects hinder the direct application of 
gradient-based optimization methods across the 
quantum-classical boundary, obstructing end-to-end 
training of hybrid models. The surrogate model 
solution, proposed in the next section, addresses 
this problem.

@@

The readout is performed by measuring specific observables,
such as $\langle Z_i \rangle$, $\langle Z_i \, Z_j\rangle$, and $\langle Z_i \, Z_j \, Z_k \rangle$, which measure respectively one-, two- and three-qubit correlations.
% \begin{itemize}
%     \item \textbf{Single-atom Measurements}: 
%     \( \langle Z_i \rangle \) for each atom \( i \), where \( Z_i \) is the Pauli-Z operator.
%     \item \textbf{Two-atom Correlations}: 
%     \( \langle Z_i Z_j \rangle \) for pairs of atoms \( (i, j) \), capturing correlations between their states.
%     \item \textbf{Three-atom Correlations}:
%     \( \langle Z_i Z_j Z_k \rangle \) for triplets of atoms \( (i, j, k) \), providing higher-order correlations.
% \end{itemize}

@@

The autoencoder was trained to minimize reconstruction 
loss using the Adam optimizer with learning rates 
between $0.001$--$0.01$ and weight decay for regularization.

@@

\subsection{Performance Metrics}
We evaluated the performance of our model using the following metrics:
Classification Accuracy, F1 Score, Confusion Matrix.

% \begin{itemize}
%     \item \textbf{Classification Accuracy} 
%     %The percentage of correctly classified instances in the test set.

%     \item \textbf{F1 Score} 
%     %The harmonic mean of precision and recall, providing a balanced measure of %classification performance.

%     \item \textbf{Confusion Matrix} 
%    % A matrix summarizing the classification results, showing true positives, %false positives, true negatives, and false negatives.

% \end{itemize}

@@

\newpage
\newpage
\tableofcontents
